---
title: "Using patchwise with locked-in and locked-out areas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using patchwise with locked-in and locked-out areas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
Suggests:
  knitr
VignetteBuilder: 
  knitr
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)

library(patchwise)
```

Prioritizations can be run using "locked-in" areas (areas that are already represented by existing protected areas) and "locked-out" areas (areas that are not available for protection). `patchwise` has the capability to include these areas while also creating the patches that should be protected as a single feature. For more information on locked-in and locked-out areas, refer to the [`prioritizr` documentation](https://prioritizr.net/).

A couple of things to note when using locked-in or locked-out areas with `patchwise`: 

1. If there are several areas you wish to lock-in, combine them into a single layer. The same goes for locked-out layers.
2. When using `patchwise` you do not need to specify locked-in or locked-out constraints via the `prioritizr` `problem()` function. `patchwise` creates the locked-in and locked-out constraints within the `features_targets()` function so that the `problem()` finds a solution with 100% of the units in the locked-in area and 0% of the units in the locked-out area.

A quick note about `prioritizr` - sometimes there are several cells that are equally worthy of protection. In this case, there may be some discrepancy in the specific areas that are selected for protection depending on the input format of the data (`raster` vs `sf`). You'll notice that this is the case in our examples here. To get more consistent results, larger protective targets and/or more features should be added to the model.

Here are some quick links to hop to different sections that you might find relevant: 

- [Implement locked-in areas using a `raster` input]
- [Implement locked-out areas using a `raster` input]
- [Implement locked-in and locked-out areas using a `raster` input]
- [Implement locked-in areas using an `sf` input]
- [Implement locked-out areas using an `sf` input]
- [Implement locked-in and locked-out areas using an `sf` input]

# Implement locked-in areas using a `raster` input

To generate our data for this example, we will use data from Bermuda generated via `oceandatr`. 

```{r}
# Choose area of interest (Bermuda EEZ)
area <- oceandatr::get_area(area_name = "Bermuda")
projection <- '+proj=laea +lon_0=-64.8108333 +lat_0=32.3571917 +datum=WGS84 +units=m +no_defs'

# Create a planning grid
planning_rast <- oceandatr::get_planning_grid(area, projection = projection)

# Grab all relevant data
features_rast <- oceandatr::get_features(planning_grid = planning_rast)

# Create a "cost" to protecting a cell - just a uniform cost for this example
cost_rast <- setNames(planning_rast, "cost")

# Separate seamount data - we want to protect entire patches
seamounts_rast <- features_rast[["seamounts"]]
features_rast <- features_rast[[names(features_rast)[names(features_rast) != "seamounts"]]]
```

We can generate a locked-in layer that would represent an existing MPA. 

```{r}
# Create a function to produce a square area
create_area <- function(planning_grid, mpa_size, edge_offset) { 
  if(class(planning_grid)[1] %in% c("RasterLayer", "SpatRaster")) { 
    terra::rast(planning_grid,
                vals = c(rep(NA, (terra::ncol(planning_grid)*edge_offset)), 
                         rep(c(rep(NA, edge_offset), 
                               rep(1, mpa_size), 
                               rep(NA, (terra::ncol(planning_grid)-(edge_offset+mpa_size)))), 
                             mpa_size), 
                         rep(NA, (terra::ncol(planning_grid)*(terra::nrow(planning_grid)-edge_offset-mpa_size)))))
  } else if(class(planning_grid)[1] == "sf") { 
    sf_to_grid <- sf::st_make_grid(planning_grid, cellsize = sqrt(sf::st_area(planning_grid[1,])))
    sf_centroid <- sf::st_centroid(sf_to_grid) %>% 
      sf::st_coordinates(.) %>% 
      as.data.frame() %>%  
      dplyr::mutate(X = round(X, digits = 4), 
                    Y = round(Y, digits = 4))
    
    # sf::st_make_grid changes the orientation of how cells are read - change them back
    sf_to_grid <- sf_to_grid %>% 
      as.data.frame() %>% 
      dplyr::bind_cols(sf_centroid) %>% 
      dplyr::arrange(dplyr::desc(Y), X) %>% 
      dplyr::select(-X, -Y)
    
    cols_sf_to_grid <- length(unique(sf_centroid$X))
    rows_sf_to_grid <- length(unique(sf_centroid$Y))
    
    sf_to_grid <- sf_to_grid %>% 
      sf::st_sf() %>%
      dplyr::mutate(layer = c(rep(NA, (cols_sf_to_grid*edge_offset)), 
                         rep(c(rep(NA, edge_offset), 
                               rep(1, mpa_size), 
                               rep(NA, (cols_sf_to_grid-(edge_offset+mpa_size)))), 
                             mpa_size), 
                         rep(NA, (cols_sf_to_grid*(rows_sf_to_grid-edge_offset-mpa_size)))))
    
    # the planning sf only includes polygons for the planning area (we need to remove them) so the number of cells matches the planning grid 
    sf_to_grid <- sf_to_grid[sf_to_grid$geometry %in% planning_grid[[1]], ]
    rownames(sf_to_grid) <- NULL
    sf_to_grid %>% 
      dplyr::relocate(layer, 1)
    }
}
```

```{r}
# Create a fake MPA
mpa_location <- create_area(planning_rast, 20, 50)

# Plot fake MPA
terra::plot(mpa_location)
```

Now we can continue using `patchwise` to group seamount areas so that entire seamount ranges are protected (and not portions of seamounts). 

```{r}
# Create seamount patches - seamount areas that touch are considered the same patch
patches_rast <- patchwise::create_patches(seamounts_rast)

# Create patches dataframe - this creates several constraints so that entire seamount units are protected together
patches_df_rast <- patchwise::create_patch_df(planning_grid = planning_rast, features = features_rast, patches = patches_rast, costs = cost_rast, locked_in = mpa_location)

# Create boundary matrix for prioritizr
boundary_matrix_rast <- patchwise::create_boundary_matrix(planning_grid = planning_rast, patches = patches_rast, patch_df = patches_df_rast)

# Create targets for protection - let's just do 20% for each feature (including 20% of whole seamounts)
targets_rast <- patchwise::features_targets(targets = rep(0.2, (terra::nlyr(features_rast) + 1)), features = features_rast, pre_patches = seamounts_rast, locked_in = mpa_location)

# Add these targets to targets for protection for the "constraints" we introduced to protect entire seamount units
constraints_rast <- patchwise::constraints_targets(feature_targets = targets_rast, patch_df = patches_df_rast)

# Run the prioritization
problem_rast <- prioritizr::problem(x = patches_df_rast, features = constraints_rast$feature, cost_column = "cost") %>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_manual_targets(constraints_rast) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_boundary_penalties(penalty = 0.000002, data = boundary_matrix_rast) %>%
  prioritizr::add_gurobi_solver(gap = 0.1, threads = parallel::detectCores()-1)

# Solve the prioritization
solution_rast <- solve(problem_rast)

# Convert the prioritization into a more digestible format
result_rast <- patchwise::convert_solution(solution = solution_rast, patch_df = patches_df_rast, planning_grid = planning_rast)

# Show the results
terra::plot(result_rast)
```

The areas in green are those selected by the prioritization for protection. We can clearly see the locked-in area (the green square) was included! 

# Implement locked-out areas using a `raster` input

What if we want to note of an area that we cannot consider for protection? We can include a "locked-out" area. We repeat the example above, but this time we will exclude this area from the prioritization results. 

```{r}
# Create a fake locked-out area
no_protection <- create_area(planning_rast, 20, 95)

# Plot fake locked-out area
terra::plot(no_protection)
```

Now we can continue using `patchwise` to group seamount areas so that entire seamount ranges are protected (and not portions of seamounts). 

```{r}
# Create patches dataframe - this creates several constraints so that entire seamount units are protected together
patches_df_rast <- patchwise::create_patch_df(planning_grid = planning_rast, features = features_rast, patches = patches_rast, costs = cost_rast, locked_out = no_protection)

# Create boundary matrix for prioritizr
boundary_matrix_rast <- patchwise::create_boundary_matrix(planning_grid = planning_rast, patches = patches_rast, patch_df = patches_df_rast)

# Create targets for protection - let's just do 20% for each feature (including 20% of whole seamounts)
targets_rast <- patchwise::features_targets(targets = rep(0.2, (terra::nlyr(features_rast) + 1)), features = features_rast, pre_patches = seamounts_rast, locked_out = no_protection)

# Add these targets to targets for protection for the "constraints" we introduced to protect entire seamount units
constraints_rast <- patchwise::constraints_targets(feature_targets = targets_rast, patch_df = patches_df_rast)

# Run the prioritization
problem_rast <- prioritizr::problem(x = patches_df_rast, features = constraints_rast$feature, cost_column = "cost") %>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_manual_targets(constraints_rast) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_boundary_penalties(penalty = 0.000002, data = boundary_matrix_rast) %>%
  prioritizr::add_gurobi_solver(gap = 0.1, threads = parallel::detectCores()-1)

# Solve the prioritization
solution_rast <- solve(problem_rast)

# Convert the prioritization into a more digestible format
result_rast <- patchwise::convert_solution(solution = solution_rast, patch_df = patches_df_rast, planning_grid = planning_rast)

# Show the results
terra::plot(result_rast)
```

Great! The area that we don't want to include in the solution is removed! 

# Implement locked-in and locked-out areas using a `raster` input

We can also run a scenario in which we have both locked-in and locked-out areas. 

```{r}
# Create patches dataframe - this creates several constraints so that entire seamount units are protected together
patches_df_rast <- patchwise::create_patch_df(planning_grid = planning_rast, features = features_rast, patches = patches_rast, costs = cost_rast, locked_in = mpa_location, locked_out = no_protection)

# Create boundary matrix for prioritizr
boundary_matrix_rast <- patchwise::create_boundary_matrix(planning_grid = planning_rast, patches = patches_rast, patch_df = patches_df_rast)

# Create targets for protection - let's just do 20% for each feature (including 20% of whole seamounts)
targets_rast <- patchwise::features_targets(targets = rep(0.2, (terra::nlyr(features_rast) + 1)), features = features_rast, pre_patches = seamounts_rast, locked_in = mpa_location, locked_out = no_protection)

# Add these targets to targets for protection for the "constraints" we introduced to protect entire seamount units
constraints_rast <- patchwise::constraints_targets(feature_targets = targets_rast, patch_df = patches_df_rast)

# Run the prioritization
problem_rast <- prioritizr::problem(x = patches_df_rast, features = constraints_rast$feature, cost_column = "cost") %>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_manual_targets(constraints_rast) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_boundary_penalties(penalty = 0.000002, data = boundary_matrix_rast) %>%
  prioritizr::add_gurobi_solver(gap = 0.1, threads = parallel::detectCores()-1)

# Solve the prioritization
solution_rast <- solve(problem_rast)

# Convert the prioritization into a more digestible format
result_rast <- patchwise::convert_solution(solution = solution_rast, patch_df = patches_df_rast, planning_grid = planning_rast)

# Show the results
terra::plot(result_rast)
```

It looks like we got the results we wanted - the locked-in areas were included in the prioritization output as an area to protect, and the locked-out areas were not included! 

# Implement locked-in areas using an `sf` input

To generate our data for this example, we will use data from Bermuda generated via `oceandatr`. 

```{r include=FALSE}
# Choose area of interest (Bermuda EEZ)
area <- oceandatr::get_area(area_name = "Bermuda")
projection <- '+proj=laea +lon_0=-64.8108333 +lat_0=32.3571917 +datum=WGS84 +units=m +no_defs'

# Create a planning grid
planning_sf <- oceandatr::get_planning_grid(area, projection = projection, option = "sf_square")

# Grab all relevant data
features_sf <- oceandatr::get_features(planning_grid = planning_sf)

# Create a "cost" to protecting a cell - just a uniform cost for this example
cost_sf <- features_sf %>%
  dplyr::mutate(cost = 1) %>%
  dplyr::select(cost)

# Separate seamount data - we want to protect entire patches
seamounts_sf <- features_sf %>% 
  dplyr::select(seamounts)

features_sf <- features_sf %>% 
  dplyr::select(-seamounts)
```

We can generate a locked-in layer that would represent an existing MPA. 

```{r}
# Create a fake MPA
mpa_location <- create_area(planning_sf, 20, 50)

# Plot fake MPA
plot(mpa_location, main = NULL, border = F)
```

Now we can continue using `patchwise` to group seamount areas so that entire seamount ranges are protected (and not portions of seamounts). 

```{r}
# Create seamount patches - seamount areas that touch are considered the same patch
patches_sf <- patchwise::create_patches(seamounts_sf, planning_grid = planning_sf)

# Create patches dataframe - this creates several constraints so that entire seamount units are protected together
patches_df_sf <- patchwise::create_patch_df(planning_grid = planning_sf, features = features_sf, patches = patches_sf, costs = cost_sf, locked_in = mpa_location)

# Create boundary matrix for prioritizr
boundary_matrix_sf <- patchwise::create_boundary_matrix(planning_grid = planning_sf, patches = patches_sf, patch_df = patches_df_sf)

# Create targets for protection - let's just do 20% for each feature (including 20% of whole seamounts)
targets_sf <- patchwise::features_targets(targets = rep(0.2, ncol(features_sf)), features = features_sf, pre_patches = seamounts_sf, locked_in = mpa_location)

# Add these targets to targets for protection for the "constraints" we introduced to protect entire seamount units
constraints_sf <- patchwise::constraints_targets(feature_targets = targets_sf, patch_df = patches_df_sf)

# Run the prioritization
problem_sf <- prioritizr::problem(x = patches_df_sf, features = constraints_sf$feature, cost_column = "cost") %>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_manual_targets(constraints_sf) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_boundary_penalties(penalty = 0.000002, data = boundary_matrix_sf) %>%
  prioritizr::add_gurobi_solver(gap = 0.1, threads = parallel::detectCores()-1)

# Solve the prioritization
solution_sf <- solve(problem_sf)

# Convert the prioritization into a more digestible format
result_sf <- patchwise::convert_solution(solution = solution_sf, patch_df = patches_df_sf, planning_grid = planning_sf)

# Show the results
plot(result_sf, main = NULL, border = F)
```

The areas in yellow are those selected by the prioritization for protection. We can clearly see the locked-in area (the yellow square) was included! 


# Implement locked-out areas using an `sf` input

What if we want to note of an area that we cannot consider for protection? We can include a "locked-out" area. We repeat the example above, but this time we will exclude this area from the prioritization results. 

```{r}
# Create a fake locked-out area
no_protection <- create_area(planning_sf, 20, 95)

# Plot fake locked-out area
plot(no_protection, main = NULL, border = F)
```

Now we can continue using `patchwise` to group seamount areas so that entire seamount ranges are protected (and not portions of seamounts). 

```{r}
# Create patches dataframe - this creates several constraints so that entire seamount units are protected together
patches_df_sf <- patchwise::create_patch_df(planning_grid = planning_sf, features = features_sf, patches = patches_sf, costs = cost_sf, locked_out = no_protection)

# Create boundary matrix for prioritizr
boundary_matrix_sf <- patchwise::create_boundary_matrix(planning_grid = planning_sf, patches = patches_sf, patch_df = patches_df_sf)

# Create targets for protection - let's just do 20% for each feature (including 20% of whole seamounts)
targets_sf <- patchwise::features_targets(targets = rep(0.2, ncol(features_sf)), features = features_sf, pre_patches = seamounts_sf, locked_out = no_protection)

# Add these targets to targets for protection for the "constraints" we introduced to protect entire seamount units
constraints_sf <- patchwise::constraints_targets(feature_targets = targets_sf, patch_df = patches_df_sf)

# Run the prioritization
problem_sf <- prioritizr::problem(x = patches_df_sf, features = constraints_sf$feature, cost_column = "cost") %>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_manual_targets(constraints_sf) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_boundary_penalties(penalty = 0.000002, data = boundary_matrix_sf) %>%
  prioritizr::add_gurobi_solver(gap = 0.1, threads = parallel::detectCores()-1)

# Solve the prioritization
solution_sf <- solve(problem_sf)

# Convert the prioritization into a more digestible format
result_sf <- patchwise::convert_solution(solution = solution_sf, patch_df = patches_df_sf, planning_grid = planning_sf)

# Show the results
plot(result_sf, main = NULL, border = F)
```
Great! The area that we don't want to include in the solution is removed!

# Implement locked-in and locked-out areas using an `sf` input

We can also run a scenario in which we have both locked-in and locked-out areas. 

```{r}
# Create patches dataframe - this creates several constraints so that entire seamount units are protected together
patches_df_sf <- patchwise::create_patch_df(planning_grid = planning_sf, features = features_sf, patches = patches_sf, costs = cost_sf, locked_in = mpa_location, locked_out = no_protection)

# Create boundary matrix for prioritizr
boundary_matrix_sf <- patchwise::create_boundary_matrix(planning_grid = planning_sf, patches = patches_sf, patch_df = patches_df_sf)

# Create targets for protection - let's just do 20% for each feature (including 20% of whole seamounts)
targets_sf <- patchwise::features_targets(targets = rep(0.2, ncol(features_sf)), features = features_sf, pre_patches = seamounts_sf, locked_in = mpa_location, locked_out = no_protection)

# Add these targets to targets for protection for the "constraints" we introduced to protect entire seamount units
constraints_sf <- patchwise::constraints_targets(feature_targets = targets_sf, patch_df = patches_df_sf)

# Run the prioritization
problem_sf <- prioritizr::problem(x = patches_df_sf, features = constraints_sf$feature, cost_column = "cost") %>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_manual_targets(constraints_sf) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_boundary_penalties(penalty = 0.000002, data = boundary_matrix_sf) %>%
  prioritizr::add_gurobi_solver(gap = 0.1, threads = parallel::detectCores()-1)

# Solve the prioritization
solution_sf <- solve(problem_sf)

# Convert the prioritization into a more digestible format
result_sf <- patchwise::convert_solution(solution = solution_sf, patch_df = patches_df_sf, planning_grid = planning_sf)

# Show the results
plot(result_sf, main = NULL, border = F)
```

It looks like we got the results we wanted - the locked-in areas were included in the prioritization output as an area to protect, and the locked-out areas were not included! 
